# Panel Data - Fixed Effects and some Random Effects

## Seminar

In this seminar, you will be asked to work more on your own. Start by clearing your workspace and setting your working directory. We will then introduce the necessary R code for today using the example from the lecture. This will be brief and afterwards, you can analyse yourself whether more guns lead to less crime.

```{r eval = FALSE, class.source="collapsible"}
rm(list = ls())
setwd("Your directory")
```



### More guns, less crime

More guns, less crime. This is the claim of an in(famous) book. It shows that violent crime rates
in the United States decrease when gun ownership restrictions are relaxed. The data used in Lott's
research compares violent crimes, robberies, and murders across 50 states to determine whether
the so called "shall" laws that remove discretion from license granting authorities actually decrease
crime rates. So far 41 states have passed these "shall" laws where a person applying for a licence
to carry a concealed weapon doesn't have to provide justification or "good cause" for requiring a
concealed weapon permit.

Load the guns.csv dataset directly into R by running the following line:

```{r}
a <- read.csv("http://philippbroniecki.github.io/philippbroniecki.github.io/assets/data/guns.csv")
```

The data includes the following variables:
* mur - Murder rate (incidents per 100,000)
* shall - =1 if state has a shall-carry law in effect in that year, 0 otherwise
* incarc rate - Incarceration rate in the state in the previous year (sentenced prisoners per 100,000 residents; value for the previous year)
* pm1029 - Percent of state population that is male, ages 10 to 29
* stateid - ID number of states (Alabama = , Alaska = 2, etc.)
* year - Year (1977 - 1999)


### Question 1

Estimate the effect of shall using a simple linear model and interpret it.

<details>
<summary>Reveal answer</summary>

```{r}
summary(lm(mur~shall+incarc_rate+pm1029,data=a))
```

> Answer: According to our simple linear model, lax gun laws reduce the murder rate. It decreases by
roughly 2 incidents per 100,000.

<details>

### Question 2 

Estimate a unit fixed effects model and a random effects model. Are both models consistent.
If not which is the appropriate model? Use a consistent model to estimate the effect of the
shall laws on the murder rate.

<details>
<summary>Reveal answer</summary>

```{r}
# panel data library
library(plm)

# fixed effects
m.fe <- plm(mur ~ shall + incarc_rate + pm1029,
data = a,
index = c("stateid", "year"),
model = "within",
effect = "individual")

# random effects
m.re <- plm(mur ~ shall + incarc_rate + pm1029,
data = a,
index = c("stateid", "year"),
model = "random")

# hausman test
phtest(m.fe, m.re)

# effect
summary(m.fe)
```

> Answer: The Hausman test shows that we reject the null hypothesis - both random effects model and
fixed effects model are consistent. The unique errors ui are correlated with the regressors.
Therefore, we must rely on the fixed effects model.

> The effect of the shall laws has decreased slightly but is still significantly related to the murder
rate. Lax gun laws reduce the murder rate by 1.45 incidents per 100,000.

<details>
<summary>Reveal answer</summary>

### Question 3

Think of a theoretical reason to control for time fixed effects (what confounding sources could
bias our estimate of the shall laws?). Test for time fixed effects using the appropriate test.
If time fixed effects are required, re-estimate the fixed effects model as a twoway fixed effects
model and interpret the effect of lax gun laws